---
title: "lnL_and_LRT"
execute:
  echo: true
  warning: false
format:
  docx:
    toc: true
    number-sections: true
    highlight-style: github
  html:
    toc: true
    number-sections: true
    self-contained: true
  pdf:
    toc: true
    number-sections: true 
editor: visual
editor_options: 
  chunk_output_type: console
---

# lnL extraction and signifcant test

## Environment setup

```{r}
#| eval: false
save.image(file = "/Users/trystan.bessonnier/Downloads/Comparative_Genomics_of_Gerromorpha/Scripts/6_CODEML/7_lnL_and_LRT.RData")
```

```{r}
#| eval: false
load(file = "/Users/trystan.bessonnier/Downloads/Comparative_Genomics_of_Gerromorpha/Scripts/6_CODEML/7_lnL_and_LRT.RData")
```

## Loading data

```{r}
setwd("/Volumes/Trystan/Lyon/M2/Stage/6_codeml")
```

```{r}
get_lnL <- function(f) {
  if (!file.exists(f)) return(NA_real_)
  x <- readLines(f, warn = FALSE)
  x <- x[grep("^lnL\\(", x)]
  if (length(x) == 0) return(NA_real_)
  m <- regexec("lnL\\(.*\\):\\s*([-0-9.]+)", x[1])
  r <- regmatches(x[1], m)[[1]]
  if (length(r) < 2) return(NA_real_)
  as.numeric(r[2])
}
```

# Site Models

## M8 (**β and ω)** vs M7 (**β)**

### Hypothesis tested :

Does adding an extra class of sites with ω \> 1 (adaptive evolution) fit the data better than a model with ten classes with flexible normalized non-synonymous ratio distribution ?

```{r}
base <- "M7M8"
ogs <- list.dirs(base, recursive = FALSE, full.names = FALSE)

res <- data.frame(OG=ogs, lnL_M7=NA_real_, lnL_M8=NA_real_)
for (i in seq_along(ogs)) {
  og <- ogs[i]
  res$lnL_M7[i] <- get_lnL(file.path(base, og, "M7.out"))
  res$lnL_M8[i] <- get_lnL(file.path(base, og, "M8.out"))
}

res$LRT <- 2*(res$lnL_M8 - res$lnL_M7)
res$pvalue <- pchisq(res$LRT, df=2, lower.tail=FALSE)
res$qvalue <- p.adjust(res$pvalue, method="BH")
res$significant <- ifelse(
  !is.na(res$qvalue) & res$qvalue < 0.05,
  "YES",
  "NO")

res <- res[order(res$qvalue, res$pvalue),]
write.table(res, file=file.path(base,"M7M8_results.tsv"),
            sep="\t", quote=FALSE, row.names=FALSE)
head(res, 20)
```

## M2a (positive selection) vs M1a (nearly neutral)

### Hypothesis tested:

Does adding a third class of sites with ω \> 1 (adaptive evolution) fit the data better than a model with two classes ω \< 1, ω = 1 ?

```{r}
base <- "M1aM2a"
ogs <- list.dirs(base, recursive = FALSE, full.names = FALSE)

res <- data.frame(OG=ogs, lnL_M1a=NA_real_, lnL_M2a=NA_real_)
for (i in seq_along(ogs)) {
  og <- ogs[i]
  res$lnL_M1a[i] <- get_lnL(file.path(base, og, "M1a.out"))
  res$lnL_M2a[i] <- get_lnL(file.path(base, og, "M2a.out"))
}

res$LRT <- 2*(res$lnL_M2a - res$lnL_M1a)
res$pvalue <- pchisq(res$LRT, df=2, lower.tail=FALSE)
res$qvalue <- p.adjust(res$pvalue, method="BH")
res$significant <- ifelse(
  !is.na(res$qvalue) & res$qvalue < 0.05,
  "YES",
  "NO")

res <- res[order(res$qvalue, res$pvalue),]
write.table(res, file=file.path(base,"M1aM2a_results.tsv"),
            sep="\t", quote=FALSE, row.names=FALSE)
head(res, 20)
```

# Branch Models

## Two-ratio vs one-ratio

### Hypothesis tested:

Are the foreground branches that you specify (Gerromorpha) more likely to have different ω from background branches (Other Heteroptera) ?

```{r}
base <- "branch_gerromorpha"
ogs <- list.dirs(base, recursive = FALSE, full.names = FALSE)

res <- data.frame(OG=ogs, lnL_one=NA_real_, lnL_two=NA_real_)
for (i in seq_along(ogs)) {
  og <- ogs[i]
  res$lnL_one[i] <- get_lnL(file.path(base, og, "one_ratio.out"))
  res$lnL_two[i] <- get_lnL(file.path(base, og, "two_ratio.out"))
}

res$LRT <- 2*(res$lnL_two - res$lnL_one)
res$pvalue <- pchisq(res$LRT, df=1, lower.tail=FALSE)
res$qvalue <- p.adjust(res$pvalue, method="BH")
res$significant <- ifelse(
  !is.na(res$qvalue) & res$qvalue < 0.05,
  "YES",
  "NO")

res <- res[order(res$qvalue, res$pvalue),]
write.table(res, file=file.path(base,"branch_gerromorpha_results.tsv"),
            sep="\t", quote=FALSE, row.names=FALSE)
head(res, 20)
```

## Free-ratio vs one-ratio

### Hypothesis tested :

For a tree of s species, is ω different among lineages ?

```{r}
base <- "free_vs_one"
ogs <- list.dirs(base, recursive = FALSE, full.names = FALSE)

res <- data.frame(OG=ogs, lnL_one=NA_real_, lnL_free=NA_real_)
for (i in seq_along(ogs)) {
  og <- ogs[i]
  res$lnL_one[i]  <- get_lnL(file.path(base, og, "one_ratio.out"))
  res$lnL_free[i] <- get_lnL(file.path(base, og, "free_ratio.out"))
}

res$LRT <- 2*(res$lnL_free - res$lnL_one)

# df = 2s-4 (s: tree of s species) :
k <- 2*21-4

res$pvalue <- pchisq(res$LRT, df=k, lower.tail=FALSE)
res$qvalue <- p.adjust(res$pvalue, method="BH")
res$significant <- ifelse(
  !is.na(res$qvalue) & res$qvalue < 0.05,
  "YES",
  "NO")

res <- res[order(res$qvalue, res$pvalue),]
write.table(res, file=file.path(base,"free_vs_one_results.tsv"),
            sep="\t", quote=FALSE, row.names=FALSE)

head(res, 20)
```

# Branch-site Models

## MA (ω \> 1) vs MA (ω=1)

### Hypothesis tested :

Is the defined “foreground branch” (Gerromorpha) more likely to contain sites with ω \> 1

## All Gerromorpha (ancestor+living species) foreground

```{r}
base <- "branchsite_gerromorpha"
ogs <- list.dirs(base, recursive = FALSE, full.names = FALSE)

res <- data.frame(OG=ogs, lnL_null=NA_real_, lnL_alt=NA_real_)
for (i in seq_along(ogs)) {
  og <- ogs[i]
  res$lnL_null[i] <- get_lnL(file.path(base, og, "bsA_null.out"))
  res$lnL_alt[i]  <- get_lnL(file.path(base, og, "bsA_alt.out"))
}

res$LRT <- 2*(res$lnL_alt - res$lnL_null)

# mixture 50:50: p = 0.5 * P(Chi1 >= LRT)
res$pvalue <- ifelse(is.na(res$LRT), NA_real_,
                     ifelse(res$LRT <= 0, 1,
                            0.5 * pchisq(res$LRT, df=1, lower.tail=FALSE)))

res$qvalue <- p.adjust(res$pvalue, method="BH")
res$significant <- ifelse(
  !is.na(res$qvalue) & res$qvalue < 0.05,
  "YES",
  "NO")

res <- res[order(res$qvalue, res$pvalue),]
write.table(res, file=file.path(base,"branchsite_gerromorpha_results.tsv"),
            sep="\t", quote=FALSE, row.names=FALSE)
head(res, 20)
```

## All ancestor Gerromorpha (ancestor only) foreground

```{r}
base <- "branchsite_ancestor_gerromorpha"
ogs <- list.dirs(base, recursive = FALSE, full.names = FALSE)

res <- data.frame(OG=ogs, lnL_null=NA_real_, lnL_alt=NA_real_)
for (i in seq_along(ogs)) {
  og <- ogs[i]
  res$lnL_null[i] <- get_lnL(file.path(base, og, "bsA_null.out"))
  res$lnL_alt[i]  <- get_lnL(file.path(base, og, "bsA_alt.out"))
}

res$LRT <- 2*(res$lnL_alt - res$lnL_null)

# mixture 50:50: p = 0.5 * P(Chi1 >= LRT)
res$pvalue <- ifelse(is.na(res$LRT), NA_real_,
                     ifelse(res$LRT <= 0, 1,
                            0.5 * pchisq(res$LRT, df=1, lower.tail=FALSE)))

res$qvalue <- p.adjust(res$pvalue, method="BH")
res$significant <- ifelse(
  !is.na(res$qvalue) & res$qvalue < 0.05,
  "YES",
  "NO")

res <- res[order(res$qvalue, res$pvalue),]
write.table(res, file=file.path(base,"branchsite_gerromorpha_results.tsv"),
            sep="\t", quote=FALSE, row.names=FALSE)
head(res, 20)
```
